The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-06-06 manuscript/5pillars.Rmd

A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A set of files under version control is called a "repository" and represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs, but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

Although DVCSs assist with code sharing, they are not considered a long-term archiving solution and
might not survive the "ten-year challenge" of code reproducibility @Perkel2020-io.
To achieve that, the code needs to be deposited to a long-term repository.

[Zenodo](https://zenodo.org/) and [Software Heritage](https://www.softwareheritage.org/) repositories
are both good options for long-term archiving of software, and the
link to these should be provided in the respective journal article/pre-print.

@@

In R, such reporting is possible using the `sessionInfo()` command,
while for Python this is possible using the `session_info` package.
Using literate programming and sharing output documents ensures that a record of this important
information is made available.

@@

Guix is described as a "functional package manager" and has the unique benefit of giving
bit-for-bit build reproducibility and verifiability [@Vallet2022-no; @Courtes2022-rx].
This is a notable advantage, as Docker images are not verifiable nor guaranteed to build reproducibly
in the future due to link decay.
Moreover, Guix can build and output environments as Docker-compatible images which solves the
provenance problem of such environments [@GNU_Project_Volunteers2022-bt].
These features are being used in the bioinformatics sphere to create highly reproducible pipelines
that work equally well on personal as well as shared high performance computers
[@Wurmus2018-lv; @Courtes2022-rx].
Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
tutorials, illustrated by the complete absence of published Guix protocols in protocols.io, as compared
to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).

>>>>

A version control system (sometimes called "source control") is a type of program that "tracks changes"
made to sets of files, typically other computer program source code, scripts and documentation.
A set of files under version control is called a "repository" and represents a project or sub-project.
Version control is used extensively by software developers and is considered one of the key best
practices in software engineering.
Distributed version control systems (DVCS) involve a central web-accessible server hosting a
repository, and each team member possesses a mirror copy on their local system (**Figure 3**).
Having a central, publicly available node assists in disseminating changes within teams and releasing
code to consumers.
There are many such DVCSs (subversion, git, bazaar, mercurial...), but `git` has emerged as the most popular
solution due to its many powerful features, speed/efficiency and large community and ecosystem.

@@

Although DVCSs assist with code sharing, they are not considered a long-term archiving solution, 
as recalled by the complete shutdown of the Google Code platform in 2015 
and the end of Mercurial support by the Bitbucket platform in 2020. 
The code needs to be deposited to a long-term repository.
[Zenodo](https://zenodo.org/) and [Figshare](https://figshare.com/) are often used 
to store the content of a repository but do not retain the history of the development of the source code.
[Software Heritage](https://www.softwareheritage.org/) @dicosmo2017 is the universal software archive 
which mission is to collect, preserve and share source code. Software Heritage provides 
an intrinsic persistent identifier called SWHID that allows to cite the archived source code 
in the respective journal article/preprint. The archival of the source code in Software Heritage
can be performed manually or automated by a webhook within a continuous integration.

@@

In R, such reporting is possible using the `sessionInfo()` command,
while for Python this is possible using the `session_info` or `watermark` packages.
Using literate programming and sharing output documents ensures that a record of this important
information is made available.

@@

Guix is described as a "functional package manager" and has the unique benefit of giving
bit-for-bit build reproducibility and verifiability [@Vallet2022-no; @Courtes2022-rx].
This is a notable advantage, as Docker images are not verifiable nor guaranteed to build reproducibly
in the future due to link decay.
Moreover, Guix can build and output environments as Docker-compatible images which solves the
provenance problem of such environments [@GNU_Project_Volunteers2022-bt].
These features are being used in the bioinformatics sphere to create highly reproducible pipelines
that work equally well on personal as well as shared high performance computers
[@Wurmus2018-lv; @Courtes2022-rx]. 
Guix is probably one of the best solution so far for the "ten-year challenge" of code reproducibility @Perkel2020-io.
Although Guix has some exciting functionality, there remains a relative lack of step-by-step guides and
tutorials, illustrated by the complete absence of published Guix protocols in protocols.io, as compared
to 13 and 12 related to "Conda" and "Docker" respectively (as at 30th May 2023).
Eventually, it's also worth mentionning that the Guix project uses Software Heritage
as a fallback system to provide a package source code
if the download of the source code from its original location failed.
